### 2012-11-19 paper/isca40format/Introduction.tex
stuff

In this paper, we present a substantially easier method to design directory-based cache-coherence protocols for such systems. The method breaks down the design problem in two parts: a part that is shared by a number of concrete protocols which we call \glob{}s, and another part that is specific to the protocol to be designed and implemented, called \policy{}. \glob{}s provide guarantees such as every request will eventually get a response, the directory's notion of the cache states is ``conservative'' \ie a cache can not have more permissions for a line than what the directory thinks it has, \etc; \glob{} enables \policy{} to focus on the high level properties such as the set of actions a cache or a directory should perform to maintain the Single-writer-at-a-time and the Read-from-last-writer invariants.

The \glob{}s can in turn be guaranteed using a set of (\policy{} independent) \local{}s which dictate local properties of a node such as when a request can be handled, what state changes happen on sending a response, properties about the network, \etc. We have formally proved that \local{}s ensure \glob{}s using the Coq Theorem Prover \cite{}, however, a presentation of the formal proof is beyond the scope of this paper. The \glob{}s and the \local{}s form the cornerstone of our framework. The separation of \glob{}s and \local{}s is akin to having an abstraction layer which hides low level details.

The \glob{}s of our framework can be used to design a wide range of cache coherence protocols, notably MSI, MESI, MOSI, MOESI, as well as their hierarchical version. We will show how these are designed as \policy{}s, focusing on high-level details.

Our method substantially reduces the design and proof burden for new protocols because of the common abstract protocol. The abstract protocol is a set of transitions that the caches and directories follow to send or receive a request or a response. We have formally proved that these transitions ensure the following properties: a) the directory's notion of a cache's state is ``conservative'' in that a cache is never going to have more permissions that what the directory thinks the cache has; and b) any request will eventually get a corresponding response. A formal proof of the abstract protocol was written using the Coq theorem prover \cite{}, however, a presentation of the formal proof is beyond the scope of this paper.

It should be noted that the flexibility of our methodology to design multiple cache coherence protocols comes from the fact that we do not assign any meaning to the value of a coherence state beyond upgrades or downgrades. Thus, how or why a particular cache state implies that a particular address may not be present in any other cache stems from protocol specific ``compatibility of state'' check.

The main contributions of this paper are: 1. A new abstract protocol that satisfies two important invariants for distributed and hierarchal message-passing protocols; 2. Instantiation of the abstract protocol to implement four well-known protocols (MSI, MESI, MOSI, MOESI) and their variations and showing why their correctness is guaranteed as long as the abstract protocol is correct, \ie it provably implements the two invariants; and 3. Our method does not introduce any overhead over direct implementation of these protocols.  We think that this new method of designing protocols substantially reduces the required effort, and that our abstract protocol is flexible enough to implement (while ensuring correctness of) most if not all inclusive, message-passing, directory-based protocols.

\paragraph{Paper organization:} Section \ref{sec:abstract} describes the abstract protocol; Section \ref{sec:properties} gives the two properties that are guaranteed by the abstract protocol and provides some intuition for proving these properties. We also discuss the communication buffer requirements for the abstract protocol and show that the requirement is only proportional to the number of levels and not to the number of processors. In Section \ref{sec:protocols}, we show how the abstract protocol can be used to implement MSI, MESI, MOSI and MOESI protocols and their variants. In Section \ref{sec:performance}, we show that these formally verified family of protocols require no extra state bits, nor extra message transfers compared to other published implementations. In Section \ref{sec:related}, we present some of the related work and finally in Section \ref{sec:conclusion}, we present some directions for the future work.

>>>>

In this paper, we present a substantially easier method to design cache
coherence protocols for such systems. The user specifies a protocol for a
two-level system without message passing. The protocol rules can simultaneously
or \emph{atomically} examine and change the state of all the caches in the
system. For example, the MSI protocol may be described as follows:

***** Murli insert the description of MSI **** We give a procedure that
transforms such simple and global protocol descriptions into a distributed,
message-passing protocol for hierarchal caches. Our procedure is based on
preserving some invariants in rule transformations and has been shown to be
sound and complete with respect the global description using the coq theorem
prover \cite{}, however, a discussion of the formal proof is beyond the scope
of this paper. Both the procedure and the proof are parameterized by the number
of cache levels and the number of caches in each level of the hierarchy as well
as by a predicate which involves a local ``compatibility of state'' test.

It should be noted that the flexibility of our procedure comes from the fact
that we do not assign any meaning to the value of a coherence state beyond
upgrades or downgrades. Thus, how or why a particular cache state implies that
a particular address may not be present in any other cache stems from protocol
specific ``compatibility of state'' check.

The main contributions of this paper are: 1. An automatic procedure to refine
centralized protocols without message passing into protocols for distributed,
hierarchal, cache system with message-passing; 2. Specification of the
invariants to be preserved by the procedure; and 3. Showing that our procedure
does not introduce any overhead over direct implementation of these protocols.
We think that our method of designing protocols substantially reduces the
required effort, and that our procedure is flexible enough to implement (while
ensuring correctness of) most, if not all, inclusive message-passing
directory-based protocols.

\paragraph{Paper organization:} *** to be redone *** Section \ref{sec:abstract}
describes the abstract protocol; Section \ref{sec:properties} gives the two
properties that are guaranteed by the abstract protocol and provides some
intuition for proving these properties. We also discuss the communication
buffer requirements for the abstract protocol and show that the requirement is
only proportional to the number of levels and not to the number of processors.
In Section \ref{sec:protocols}, we show how the abstract protocol can be used
to implement MSI, MESI, MOSI and MOESI protocols and their variants.  In
Section \ref{sec:performance}, we show that these formally verified family of
protocols require no extra state bits, nor extra message transfers compared to
other published implementations.  In Section \ref{sec:related}, we present some
of the related work and finally in Section \ref{sec:conclusion}, we present
some directions for the future work.

### 2012-11-19 paper/isca40format/Invariants.tex
stuff

\begin{figure}
\begin{inv}
An incoming request message from a child should not block an incoming request
message from the parent, even for a different address\label{cReqNoBlockPReq}
\end{inv}
\begin{inv}
An incoming request messages for an address $a$ should not block an incoming
response message for address $a$, even from a different
source.\label{reqNoBlockReq}
\end{inv}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

\begin{figure}
\begin{inv}
An incoming response message for an address $a$ from a child $c$ should not be
received before another incoming response message from the same $c$ and the
same $a$ sent earlier has been received\label{cRespFifo}
\end{inv}
\begin{inv}
An incoming request message for an address $a$ from a source $n$ should not be
received before another incoming response message from the same $n$ and the
same $a$ sent earlier has been received\label{reqNoOvertakeResp}
\end{inv}
\caption{Ordering requirements}
\label{order}
\end{figure}

\begin{figure}
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a thread that handling a request from $c$ to $p$ for
address $a$ is executing, and
\item $x > p.dir[c][a]$
\end{enumerate}
\label{cSendRespPre}
\end{inv}
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
\label{cSendRespPost}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < c.state[a]$ \label{pSendRespPre1}
\end{inv}
\begin{inv}
If there is a thread in node $c$ that is waiting to receive a response from its
parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
on executing a thread that is handling a request from $p$ to $c$ for address
$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
handling a request from $p$ to $c$ for address $a$.
\label{pSendRespPre2}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
\label{pSendRespPost}
\end{inv}
\caption{Invariants for sending responses}
\label{sendResp}
\end{figure}

\begin{figure}
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$
\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$
\label{cRecvResp}
\end{inv}
\caption{Invariants for receiving responses}
\label{recvResp}
\end{figure}

\begin{figure}
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$
\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$
\label{pState}
\end{inv}
\caption{Invariants for state changes}
\label{stateChange}
\end{figure}

\begin{figure}
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
\label{pSendReqPre}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
\label{cSendReqPre}
\end{inv}
\caption{Invariants for sending a request}
\label{sendReq}
\end{figure}

\begin{figure}
\begin{inv}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{inv}
\begin{inv}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from $p$, or
\item a thread in $p$ is handling a request for address
$a$ from one of $p$'s children $c'$, and that thread will not send any
(downgrade) request to any child for address $a$ anymore
\end{enumerate}
\label{cHandleReq}
\end{inv}
\caption{Invariants for receiving a request (used by the scheduler)}
\label{recvReq}
\end{figure}

\begin{figure}
\begin{inv}
Address $a$ present in node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{evict}
\caption{Invariants for suspending a thread during choosing an address for eviction}
\end{inv}
\end{figure}

\begin{figure}
\begin{inv}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled by the scheduler to execute.
\end{inv}
\end{figure}

>>>>

\begin{figure}\small
\begin{inv}
An incoming request message from a child or the core should not block an
incoming request message from the parent, even for a different
address\label{cReqNoBlockPReq}
\end{inv}
\begin{inv}
An incoming request messages for an address $a$ should not block an incoming
response message for address $a$, even from a different
source.\label{reqNoBlockResp}
\end{inv}
\caption{Blocking restrictions}
\label{blocking}
\end{figure}

\begin{figure}\small
\begin{inv}
An incoming response message for an address $a$ from a child $c$ should not be
received before another incoming response message from the same $c$ and the
same $a$ sent earlier has been received\label{cRespFifo}
\end{inv}
\begin{inv}
An incoming request message for an address $a$ from a source $n$ should not be
received before another incoming response message from the same $n$ and the
same $a$ sent earlier has been received\label{reqNoOvertakeResp}
\end{inv}
\caption{Ordering requirements}
\label{order}
\end{figure}

\begin{figure}\small
\begin{inv}
A response \Resp{p}{c}{a}{x}, where $p = c.parent$, can be sent from $p$ only
if
\begin{enumerate}
\item a thread that handling a request from $c$ to $p$ for
address $a$ is executing, and
\item $x > p.dir[c][a]$
\end{enumerate}
\label{cSendRespPre}
\end{inv}
\begin{inv}
If a response \Resp{p}{c}{a}{x} is sent from $p$, then $p.dir[c][a] \gets x$
\label{cSendRespPost}
\end{inv}
\begin{inv}
A response \Resp{c}{p}{a}{x}, where $p = c.parent$, can be sent from $c$ only
if $x < c.state[a]$ \label{pSendRespPre1}
\end{inv}
\begin{inv}
If there is a thread in node $c$ that is waiting to receive a response from its
parent $p$ for an address $a$, then response \Resp{c}{p}{a}{x} can be sent only
on executing a thread that is handling a request from $p$ to $c$ for address
$a$. For instance, $a$ can not be evicted from $c$, if there is a thread
handling a request from $p$ to $c$ for address $a$.
\label{pSendRespPre2}
\end{inv}
\begin{inv}
If \Resp{c}{p}{a}{x} is sent $c.state[a] \gets x$
\label{pSendRespPost}
\end{inv}
\caption{Invariants for sending responses}
\label{sendResp}
\end{figure}

\begin{figure}\small
\begin{inv}
On receiving a response \Resp{c}{p}{a}{x}, where $p = c.parent$, $p.dir[c][a]
\gets x$
\label{pRecvResp}
\end{inv}
\begin{inv}
On receiving a response \Resp{p}{c}{a}{x}, where $p = c.parent$, $c.state[c]
\gets x$
\label{cRecvResp}
\end{inv}
\caption{Invariants for receiving responses}
\label{recvResp}
\end{figure}

\begin{figure}\small
\begin{inv}
$c.state[a]$ can change only on $c$ sending a response to or receiving a
response from $c.parent$
\label{cState}
\end{inv}
\begin{inv}
$p.dir[c][a]$ can change only on $p$ sending or receiving a response from
its child $c$
\label{pState}
\end{inv}
\caption{Invariants for state changes}
\label{stateChange}
\end{figure}

\begin{figure}\small
\begin{inv}
\Req{p}{c}{a}{x}, where $p = c.parent$, can be sent only if $x < p.dir[c][a]$
\label{pSendReqPre}
\end{inv}
\begin{inv}
\Req{c}{p}{a}{x}, where $p = c.parent$, can be sent only if $x > c.state[a]$
\label{cSendReqPre}
\end{inv}
\caption{Invariants for sending a request}
\label{sendReq}
\end{figure}

\begin{figure}\small
\begin{inv}
$p$ can not receive a request \Req{c}{p}{a}{x} from $c$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{pHandleReq}
\end{inv}
\begin{inv}
$c$ can not receive a request \Req{p}{c}{a}{x} from $p$, where $p = c.parent$,
if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $p$ is handling a request for address
$a$ from $p$, or
\item a thread in $p$ is handling a request for address
$a$ from one of $p$'s children $c'$, and that thread will not send any
(downgrade) request to any child for address $a$ anymore
\end{enumerate}
\label{cHandleReq}
\end{inv}
\caption{Invariants for receiving a request (used by the scheduler)}
\label{recvReq}
\end{figure}

\begin{figure}\small
\begin{inv}
Address $a$ present in node $c$ can not be chosen for eviction by a thread
executing in $c$ if either
\begin{enumerate}
\item address $a$ has been chosen for eviction by some thread and that thread
has not yet replaced $a$, or
\item a thread in $c$ is handling a request for address
$a$ from any source
\end{enumerate}
\label{evict}
\caption{Invariants for suspending a thread during choosing an address for eviction}
\end{inv}
\end{figure}

\begin{figure}\small
\begin{inv}
Each node has 
\begin{enumerate}
\item at least one dedicated thread resource for handling only messages (requests or
responses) from the parent,
\item at least one dedicated thread resource for handling only requests from any
child, and
\item at least one dedicated thread resource for handling only responses (from children
or from the parent).
\end{enumerate}
\label{dedicate}
\end{inv}
\caption{Dedicated resources in a node}
\end{figure}

\begin{figure}\small
\begin{inv}
Starvation-freedom: A thread that keeps becoming ready to execute will
eventually be scheduled to execute.
\end{inv}
\end{figure}

### 2012-11-19 paper/isca40format/Theorems.tex
stuff

@@

These theorems are needed to guarantee that Figure \ref{realistic} protocol is
indeed correct. If, for instance, a thread wants to send a message in the output
channel of the node and is not able to send one forever, then the thread can not
proceed further. Similarly, if a thread is waiting for a response (for a request
previously sent), and it never gets the response, the thread can never proceed
further. These scenarios create a deadlock.

>>>>

\begin{theorem}
If a thread is waiting for a line to be allocated, \ie for any address to become
available for eviction, then some address will eventually become available for
eviction
\label{evictDead}
\end{theorem}

@@

These theorems are needed to guarantee that Figure \ref{realistic} protocol is
indeed correct. If, for instance, a thread wants to send a message in the
output channel of the node and is not able to send one forever, then the thread
can not proceed further. If a thread is waiting for a response (for a request
previously sent), and it never gets the response, the thread can never proceed
further. A thread can also not proceed if it's waiting for being allocated in a
free line but never gets one because all the lines are locked up. These
scenarios create a deadlock.
